---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.path = "man/figures/README-"
)
set.seed(111)
```

# _{{ packagename }}_: Simulate infectious disease transmission with contact tracing

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/{{ gh_repo }})
[![R-CMD-check](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/{{ gh_repo }}/branch/main/graph/badge.svg)](https://app.codecov.io/gh/{{ gh_repo }}?branch=main)
![GitHub contributors](https://img.shields.io/github/contributors/{{ gh_repo }})
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

_{{ packagename }}_ is an R package that provides methods to simulate infectious disease transmission in the presence of contact tracing.
It was initially developed to support a paper written in early 2020 to assess the [feasibility of controlling COVID-19](https://github.com/cmmid/ringbp).
For more details on the methods implemented here, see the associated [paper](https://doi.org/10.1016/S2214-109X(20)30074-7).

## Installation

The current development version of _{{ packagename }}_ can be installed from [GitHub](https://github.com/) using the `pak` package.

```r
if(!require("pak")) install.packages("pak")
pak::pak("{{ gh_repo }}")
```

## Quick start

The main functionality of the package is in the `scenario_sim()` function.
Here is an example for running 10 simulations of a given scenario:

```{r scenario_sim}
library("ringbp")
library("ggplot2")

res <- scenario_sim(
  n = 10, ## 10 simulations
  initial_cases = 1, ## one initial case in each of the simulations 
  offspring = offspring_opts(
    ## non-isolated individuals have R0 of 2.5 and a dispersion parameter
    community = \(n) rnbinom(n = n, mu = 2.5, size = 0.16), 
    ## isolated individuals have R0 of 0.5 and a dispersion  parameter
    isolated = \(n) rnbinom(n = n, mu = 0.5, size = 1)
    ## by default asymptomatic individuals are assumed to have the same R0 
    ## and dispersion as non-isolated individuals
  ), 
  delays = delay_opts(
    incubation_period = \(x) stats::rweibull(n = x, shape = 2.322737, scale = 6.492272), 
    onset_to_isolation = \(x) stats::rweibull(n = x, shape = 1.651524, scale = 4.287786)
  ),
  event_probs = event_prob_opts(
    ## 10% asymptomatic infections
    asymptomatic = 0.1, 
    ## probability of onward infection time being before symptom onset
    presymptomatic_transmission = 0.5,
    ## 20% probability of ascertainment by contact tracing
    symptomatic_ascertained = 0.2
  ),
  ## whether quarantine is in effect
  interventions = intervention_opts(quarantine = FALSE),
  sim = sim_opts(
    ## don't simulate beyond 350 days
    cap_max_days = 350, 
    ## don't simulate beyond 4500 infections
    cap_cases = 4500
  )
)
```

### Plot of weekly cases

```{r plot}
ggplot(
  data = res, aes(x = week, y = cumulative, col = as.factor(sim))
) +
  geom_line(show.legend = FALSE, alpha = 0.3) +
  scale_y_continuous(name = "Cumulative number of cases") +
  theme_bw()
```

### Estimate extinction probability

```{r extinct_prob}
extinct_prob(res, cap_cases = 4500)
```

## Contributors


<!-- ALL-CONTRIBUTORS-LIST:START - Do not remove or modify this section -->
<!-- prettier-ignore-start -->
<!-- markdownlint-disable -->

All contributions to this project are gratefully acknowledged using the [`allcontributors` package](https://github.com/ropensci/allcontributors) following the [all-contributors](https://allcontributors.org) specification. Contributions of any kind are welcome!

### Code


<a href="https://github.com/epiforecasts/ringbp/commits?author=seabbs">seabbs</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=sbfnk">sbfnk</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=jhellewell14">jhellewell14</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=timcdlucas">timcdlucas</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=amygimma">amygimma</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=joshwlambert">joshwlambert</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=Bisaloo">Bisaloo</a>, 
<a href="https://github.com/epiforecasts/ringbp/commits?author=actions-user">actions-user</a>



### Issue Authors


<a href="https://github.com/epiforecasts/ringbp/issues?q=is%3Aissue+author%3Apearsonca">pearsonca</a>, 
<a href="https://github.com/epiforecasts/ringbp/issues?q=is%3Aissue+author%3Asophiemeakin">sophiemeakin</a>



### Issue Contributors


<a href="https://github.com/epiforecasts/ringbp/issues?q=is%3Aissue+commenter%3Athimotei">thimotei</a>, 
<a href="https://github.com/epiforecasts/ringbp/issues?q=is%3Aissue+commenter%3Aadamkucharski">adamkucharski</a>, 
<a href="https://github.com/epiforecasts/ringbp/issues?q=is%3Aissue+commenter%3Adcadam">dcadam</a>, 
<a href="https://github.com/epiforecasts/ringbp/issues?q=is%3Aissue+commenter%3Ajamesmbaazam">jamesmbaazam</a>


<!-- markdownlint-enable -->
<!-- prettier-ignore-end -->
<!-- ALL-CONTRIBUTORS-LIST:END -->

