---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.path = "man/figures/README-"
)
set.seed(111)
```

# _{{ packagename }}_: Simulate infectious disease transmission with contact tracing

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/{{ gh_repo }})
[![R-CMD-check](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/{{ gh_repo }}/branch/main/graph/badge.svg)](https://app.codecov.io/gh/{{ gh_repo }}?branch=main)
![GitHub contributors](https://img.shields.io/github/contributors/{{ gh_repo }})
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

_{{ packagename }}_ is an R package that provides methods to simulate infectious disease transmission in the presence of contact tracing.
It was initially developed to support a paper written in early 2020 to assess the [feasibility of controlling COVID-19](https://github.com/cmmid/ringbp).
For more details on the methods implemented here, see the associated [paper](https://doi.org/10.1016/S2214-109X(20)30074-7).

## Installation

The current development version of _{{ packagename }}_ can be installed from [GitHub](https://github.com/) using the `pak` package.

```r
if(!require("pak")) install.packages("pak")
pak::pak("{{ gh_repo }}")
```

## Quick start

The main functionality of the package is in the `scenario_sim()` function.
Here is an example for running 10 simulations of a given scenario:

```{r scenario_sim}
library("ringbp")
library("ggplot2")

res <- scenario_sim(
  n.sim = 10, ## 10 simulations
  num.initial.cases = 1, ## one initial case in each of the simulations
  prop.asym = 0, ## no asymptomatic infections
  prop.ascertain = 0.2, ## 20% probability of ascertainment by contact tracing
  cap_cases = 4500, ## don't simulate beyond 4500 infections
  cap_max_days = 350, ## don't simulate beyond 350 days
  r0isolated = 0.5, ## isolated individuals have R0 of 0.5
  r0community = 2.5, ## non-isolated individuals have R0 of 2.5
  disp.com = 0.16, ## dispersion parameter in the community
  disp.iso = 1, ## dispersion  parameter of those isolated
  onset_to_isolation = \(x) stats::rweibull(n = x, shape = 1.651524, scale = 4.287786), ## time from onset to isolation
  incfn = \(x) stats::rweibull(n = x, shape = 2.322737, scale = 6.492272),
  k = 0, ## skew of generation interval to be beyond onset of symptoms
  quarantine = FALSE ## whether quarantine is in effect
)
```

### Plot of weekly cases

```{r plot}
ggplot(
  data = res, aes(x = week, y = cumulative, col = as.factor(sim))
) +
  geom_line(show.legend = FALSE, alpha = 0.3) +
  scale_y_continuous(name = "Cumulative number of cases") +
  theme_bw()
```

### Estimate extinction probability

```{r extinct_prob}
extinct_prob(res, cap_cases = 4500)
```
