---
title: "Scenario parameter sweep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scenario parameter sweep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to run an analysis using the {ringbp} simulation across multiple scenarios.

::: {.alert .alert-info}
The functionality included in this vignette was previously in the `parameter_sweep()` function in {ringbp} <= v0.1.2. This function has been removed as it is preferred to demonstrate the functionality but allow users the flexibility and control to run scenarios in whichever way they choose.
:::

```{r setup}
library(ringbp)
library(data.table)
```

The `scenario_sim()` function is the core function to run outbreak simulations using {ringbp}, because it allows running the `outbreak_model()` over multiple replicates. 

## Set up scenario parameter space

The first step to construct a `data.frame` (in this example we will use a `data.table`) with a grid of parameters, i.e. a parameter space to explore. We use `expand.grid()` to create all the combinations of parameters.

Parameters that are grouped are put into a nested `data.table` to make sure they are coupled and combinations of these parameters are not mixed. See `delay_group` below for an example of this.

```{r}
# Put parameters that are grouped by disease into this data.table
scenarios <- data.table(
  expand.grid(
   delay_group = list(data.table(
     delay = c("SARS", "Wuhan"),
     onset_to_isolation = c(
       \(n) rweibull(n = n, shape = 1.651524, scale = 4.287786),
       \(n) rweibull(n = n, shape = 2.305172, scale = 9.483875)
     )
   )),
   r0_community = c(1.1, 1.5),
   r0_isolated = 0,
   disp_community = 0.16,
   disp_isolated = 1,
   prop_presymptomatic = c(0.01, 0.15),
   prop_asymptomatic = c(0, 0.1),
   prop_ascertain = seq(0, 1, 0.25),
   initial_cases = c(5, 10),
   quarantine = FALSE,
   cap_max_days = 365,
   cap_cases = 5000
 )
)
```

To unnest the `data.table` in order to sweep across the scenarios, a few data manipulation steps are required:

```{r}
list_cols <- grep("_group", colnames(scenarios), value = TRUE)
non_list_cols <- setdiff(colnames(scenarios), list_cols)
scenarios <- scenarios[, rbindlist(delay_group), by = c(non_list_cols)]
```

Lastly in setting up the parameter grid we add a column called `scenario` which is numbered 1 to the total number of scenarios; and we add the incubation period to the parameter grid.

```{r}
scenarios[, scenario :=  1:.N]
incub <- \(n) rweibull(n = n, shape = 1.65, scale = 4.28)
scenarios[, incubation_period := rep(list(incub), .N)]
```

## Parameter sweep across scenarios

Before we run `scenario_sim()` across each parameter set, we split the scenarios into a list. This makes it easy to loop over the resulting list using R functions like `lapply()`. 

```{r}
scenario_sims <- scenarios[, list(data = list(.SD)), by = scenario]
```

For this example we will run 3 replicates for each parameter set. The simulation model is stochastic so by running multiple replicates for each, we can understand the variance within scenarios. For a scientifically robust analysis, the number of replicates should be much higher (e.g. 100-1000).

```{r}
n <- 3
res <- lapply(scenario_sims$data, \(x, n) {
  scenario_sim(
    n = n,
    initial_cases = x$initial_cases,
    offspring = offspring_opts(
      community = \(n) rnbinom(n = n, mu = x$r0_community, size = x$disp_community),
      isolated = \(n) rnbinom(n = n, mu = x$r0_isolated, size = x$disp_isolated)
    ),
    delays = delay_opts(
      incubation_period = x$incubation_period[[1]],
      onset_to_isolation = x$onset_to_isolation[[1]]
    ),
    event_probs = event_prob_opts(
      asymptomatic = x$prop_asymptomatic,
      presymptomatic_transmission = x$prop_presymptomatic,
      symptomatic_ascertained = x$prop_ascertain
    ),
    interventions = intervention_opts(quarantine = x$quarantine),
    sim = sim_opts(cap_max_days = x$cap_max_days, cap_cases = x$cap_cases)
  )
  },
  n = n
)
```

The output of this parameter sweep is a list of `data.table`s, so to give an idea of what the output looks like we show the first 3 scenarios (there are `r length(res)` in total). See `?scenario_sim` for more information on the contents of these `data.table`s.

```{r}
head(res, 3)
```

## Storing simulations with scenarios

In the previous example we constructed the parameter grid and then ran the parameter sweep storing the simulated outbreaks in a separate object. It can be useful to keep the simulated output together with the parameter grid to be able to check which parameters correspond to which simulated outbreak.

In this example we run the same scenario parameter sweep as before (again with `r n` replicates), but store the simulated outbreak `data.table`s in the same object as the list of scenarios. We append the simulation results to the `scenario_sims` `data.table` and assign them to the `sims` column.

```{r}
scenario_sims[, sims := lapply(data, \(x, n) {
  scenario_sim(
    n = n,
    initial_cases = x$initial_cases,
    offspring = offspring_opts(
      community = \(n) rnbinom(n = n, mu = x$r0_community, size = x$disp_community),
      isolated = \(n) rnbinom(n = n, mu = x$r0_isolated, size = x$disp_isolated)
    ),
    delays = delay_opts(
      incubation_period = x$incubation_period[[1]],
      onset_to_isolation = x$onset_to_isolation[[1]]
    ),
    event_probs = event_prob_opts(
      asymptomatic = x$prop_asymptomatic,
      presymptomatic_transmission = x$prop_presymptomatic,
      symptomatic_ascertained = x$prop_ascertain
    ),
    interventions = intervention_opts(quarantine = x$quarantine),
    sim = sim_opts(cap_max_days = x$cap_max_days, cap_cases = x$cap_cases)
  )
  },
  n = n
)]
scenario_sims
head(scenario_sims$data, 3)
head(scenario_sims$sims, 3)
```

## Running scenarios in parallel

Finally, we demonstrate how the above example can be easily run in parallel using the [{future} framework](https://www.futureverse.org/). We load the {future} and {future.apply} R packages for this.

```{r}
library(future)
library(future.apply)
```

We can replace the `lapply()` with the `future.apply::future_lapply()` function and then choose how to run the analysis in parallel. See `?future::plan` for details.

Here we show how to run the analysis on 2 separate R sessions running in the background.

```{r, eval=FALSE}
future::plan("multisession", workers = 2)
```

Now we re-run the parameter sweep with parallelisation.

```{r, eval=FALSE}
scenario_sims[, sims := future_lapply(data, \(x, n) {
  scenario_sim(
    n = n,
    initial_cases = x$initial_cases,
    offspring = offspring_opts(
      community = \(n) rnbinom(n = n, mu = x$r0_community, size = x$disp_community),
      isolated = \(n) rnbinom(n = n, mu = x$r0_isolated, size = x$disp_isolated)
    ),
    delays = delay_opts(
      incubation_period = x$incubation_period[[1]],
      onset_to_isolation = x$onset_to_isolation[[1]]
    ),
    event_probs = event_prob_opts(
      asymptomatic = x$prop_asymptomatic,
      presymptomatic_transmission = x$prop_presymptomatic,
      symptomatic_ascertained = x$prop_ascertain
    ),
    interventions = intervention_opts(quarantine = x$quarantine),
    sim = sim_opts(cap_max_days = x$cap_max_days, cap_cases = x$cap_cases)
  )
  },
  n = n,
  future.seed = TRUE
)]
scenario_sims
```

Note, we set the `future.seed = TRUE` to ensure that parallel-safe random numbers are produced.
